---
title: "ESM 244 - HW4"
author: "Lauren Puffer"
format: html
editor: visual
embed-resources: true
code-fold: show
execute: 
  warning: false
  message: false
toc: TRUE
theme: journal
---

# Overview

### Purpose of Analysis

With this analysis, we aim to assess possible temporal trends of Coho salmon, Jack Coho salmon, and Steelhead trout in the Columbia River Basin. Our analysis will use a simple data manipulation in the 'lubridate' package for compatibility with visualization tools. We will then explore patterns in seasonality and overall trends, as well as compare fish counts by month using 'ggsubseries.'

### *insert photo here*

### Data citation

Data were shared by and accessed from [Columbia River DART](http://www.cbr.washington.edu/dart/query/adult_graph_text) (Data Access in Real Time), accessed January 25, 2023. This data comes from a collaboration of 8 entities monitoring fish populations in the Columbia Basin. Fish counts for 9 species and water temperature is recorded daily.

### Pseudocode

1.  Load necessary packages, including fable and feast
2.  Read in the data and use 'lubridate' to transform to month-year format
3.  Put data into a tsibble table for ease of use
4.  Conduct exploratory data analysis to look for obvious trends
5.  Create visualizations to analyze possible seasonality and trends in time series data

## Load packages

```{r}
library(tidyverse)
library(janitor)
library(here)
library(dplyr) #change NAs to zeros
library(tsibble) #make a dataframe
library(lubridate) #used for altering time series data
library(feasts) 
library(fable)
```

## Data wrangling

```{r load data and prepare for time series analysis}
# load data from local file
fish_data <- read.csv(here("data", "willamette_fish_passage.csv")) |>
  clean_names() |>
  mutate_all(~replace(., is.na(.), 0))|>  #change NA values to zeros 
  select(date, steelhead, coho, jack_coho)|>
  mutate(date = mdy(date))


#create a distinct dataframe with species in 1 column and fish count in another column  
fish_data_long <- fish_data|> 
  pivot_longer(cols = c(jack_coho, coho, steelhead), 
               names_to= "type", 
               values_to = "count")

# create a tsibbble table ofr usability 
fish_ts <- fish_data_long |>
  as_tsibble(key = type, #not more than one variable dertmining time indeces
             index = date) #use date to specify time

```

## Exploratory Data Analysis

```{r plot data}
#| fig-cap: "Fig. 1 - Daily fish counts for adult Steelhead trout, jack Coho salmon, and Coho salmon observed in the Columbia River Basin."

ggplot(data = fish_ts, 
       aes(x = date, y = count, colour = type)) +
  geom_line() +
  labs(x = "Date", 
       y = "Daily Counts",
       title = "Daily Counts by Type") +
  scale_color_manual(values = c("steelhead" = "goldenrod1", 
                                "jack_coho" = "deeppink3", 
                                "coho" = "chartreuse3")) +  # Custom colors
  theme_classic()

```

Based on the information shown in figure 1, it seems that the amount of Coho salmon in the basin spiked in the late 2000s and early 2010s. The seasonality for Stee;head and Coho salmon seem to alternate.

geom_stream

::: panel-tabset
### title

```{r}
ggplot(data = fish_ts, aes(x = date, y = count)) +
  geom_line() +
  labs(x = "Date",
       y = "Mean daily air temperature (Celsius)\n at Toolik Station")
```
:::
